Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-17 latex/main.tex
Minor tweaking on the Handling outlier samples section

Although we expected that most samples can be explained by a small number of mutations in the ARG,
we observed two types of outlier samples which had unusually high numbers of mutations:
(1) ``time travelers'' and (2) ``saltation'' lineages.
Time traveling samples are caused by errors in their collection dates.
For example, if a sample is actually collected 10 months after its reported collection date,
then 20 mutations (instead of, say, 0 or 1) are needed to explain it
using the ARG available on its collection date,
assuming a substitution rate of 2 events per genome per month.
Samples from ``saltation'' lineages (particularly the major VOCs) are characterized
by high numbers of mutations by the time of their detection.
Both types of outlier samples are poorly fit to the ARG when they appear on their collection dates.
To handle these outlier samples, we introduce three mechanisms:
(1) HMM cost-based rejection,
(2) retrospective sample grouping, and
(3) unconditional seeding.

>>>>

Although most samples can be explained by a small number of mutations in the ARG,
there are two types of outlier samples which have unusually high numbers of mutations:
``time travelers'' and ``saltation'' lineages.
Time traveling samples are caused by errors in their collection dates.
For example, if a sample is actually collected 10 months after its reported collection date,
then 20 mutations (instead of, say, 0 or 1) are needed to explain it
using the ARG available on its collection date,
assuming a substitution rate of 2 events per genome per month.
Samples from ``saltation'' lineages (particularly the major VOCs) are characterized
by high numbers of mutations by the time of their detection.
Both types of outlier samples are poorly fit to the ARG when they appear on their collection dates.
To handle these outlier samples, we introduce three mechanisms:
(1) HMM cost-based rejection,
(2) retrospective sample grouping, and
(3) unconditional seeding.
